---
title: "Tanzania Assignment 2"
author: 'Group: Alona Sycheska'
output:
  pdf_document:
    toc: true
    toc_depth: '3'
  html_document:
    toc: true
    toc_depth: 3
    df_print: paged
    number_sections: true
---

```{r setup, include=FALSE}
library(haven)
library(dplyr)
library(tidyr)
library(knitr)
library(broom)
library(ggplot2)
```

# 1. Load Data

```{r}
d18 <- read_dta("Tanzania_2018_new.dta")
glimpse(d18)
```

# 2. Poverty Indicator

```{r}
d18 <- d18 %>% mutate(poor_nat = if_else(cons < povline, 1L, 0L))
table(d18$poor_nat)
```

**Answer:**  
A person is considered poor if `cons < povline`. This matches the national poverty line definition.

# 3. Percent of Children in School (2018)

```{r}
d18_children <- d18 %>%
  mutate(
    loc = if_else(rural == 1, "Rural", "Urban"),
    sex = if_else(male == 1, "Male", "Female"),
    age_group = case_when(
      age >= 7 & age <= 12 ~ "7-12",
      age >= 13 & age <= 18 ~ "13-18",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(poor_nat == 1, !is.na(age_group), !is.na(inschool))

tab_2018 <- d18_children %>%
  group_by(loc, sex, age_group) %>%
  summarise(enrol_2018 = 100 * weighted.mean(inschool == 1, w = hhweight, na.rm = TRUE))

kable(tab_2018)
```
## Graph 1: Enrollment Rates in 2018 by Group

```{r}
ggplot(tab_2018, aes(x=sex, y=enrol_2018, fill=loc)) +
  geom_col(position="dodge") +
  facet_wrap(~age_group) +
  labs(title="School Enrollment of Poor Children in Tanzania (2018)",
       y="Enrollment Rate (%)",
       x="Sex",
       fill="Location") +
  theme_minimal()
```

# 4. Comparison with Banerjee & Duflo (Table 9)

```{r}
tab_bd <- tibble::tribble(
  ~loc, ~sex, ~age_group, ~enrol_2004,
  "Rural","Female","7-12",51.2,
  "Rural","Male","7-12",47.2,
  "Urban","Female","7-12",66.4,
  "Urban","Male","7-12",54.4,
  "Rural","Female","13-18",53.3,
  "Rural","Male","13-18",61.4,
  "Urban","Female","13-18",51.8,
  "Urban","Male","13-18",65.3
)

comp <- inner_join(tab_bd, tab_2018, by=c("loc","sex","age_group")) %>%
  mutate(change_pp = enrol_2018 - enrol_2004)

kable(comp)
```

## Graph 2: Change Over Time (2004 vs 2018)

```{r}
comp_long <- comp %>%
  pivot_longer(cols=c(enrol_2004, enrol_2018), names_to="year", values_to="value")

ggplot(comp_long, aes(x=sex, y=value, fill=year)) +
  geom_col(position="dodge") +
  facet_grid(loc ~ age_group) +
  labs(title="School Enrollment: 2004 vs 2018",
       x="Sex",
       y="Enrollment (%)",
       fill="Year") +
  theme_minimal()
```

# Policy Note

**Background:**  
Enrollment among poor Tanzanian children has changed between early-2000s and 2018. The goal is to understand improvements and remaining gaps.

**Key Findings:**  
- Primary school (7–12) enrollment increased for all groups.  
- Rural children still lag behind urban children, but the gap narrowed.  
- Gender differences decreased, with females showing strong improvements.  
- Secondary school enrollment (13–18) improved but remains lower than primary.

**Interpretation:**  
- Gains reflect national policies boosting primary education.  
- Rural improvements may come from infrastructure investments.  
- Secondary education barriers persist: costs, distance, work responsibilities.

**Policy Recommendations:**  
- Expand support for secondary education (books, uniforms, transport).  
- Provide cash transfers for rural poor households.  
- Support adolescent girls via targeted programs.

# 5. RCT: Free Books

```{r}
rct <- d18 %>% 
  filter(age >= 7, age <= 12) %>%
  mutate(
    loc = factor(if_else(rural == 1, "Rural","Urban")),
    sex = factor(if_else(male == 1, "Male","Female"))
  )

rct_tab <- rct %>%
  group_by(freebooks) %>%
  summarise(enrol_later = 100 * weighted.mean(inschool_later == 1, w = hhweight, na.rm = TRUE))

kable(rct_tab)
```

## Graph 3: RCT Treatment vs Control Enrollment

```{r}
ggplot(rct_tab, aes(x=factor(freebooks), y=enrol_later, fill=factor(freebooks))) +
  geom_col() +
  scale_x_discrete(labels=c("0"="Control","1"="Free Books")) +
  labs(title="Impact of Free Books on Later Enrollment",
       x="Group",
       y="Enrollment (%)",
       fill="Group") +
  theme_minimal()
```

# 6. Regression Analysis

## 6.1 Main Treatment Effect

```{r}
lm1 <- lm(inschool_later ~ freebooks, data=rct, weights=hhweight)
kable(tidy(lm1))
```

### Interpretation  
The intercept (0.902) means about 90.2% of children in the control group were enrolled in school later.

The coefficient on freebooks = 0.0098 indicates:

Children who received free books are 0.98 percentage points more likely to be enrolled later.

The p-value (0.035) shows this effect is statistically significant at the 5% level.


While the effect is statistically significant, it is small in magnitude.
Free books helped, but the improvement is modest — under 1 percentage point.

## 6.2 Controlled Regression

```{r}
lm2 <- lm(inschool_later ~ freebooks + age + sex + loc, data=rct, weights=hhweight)
kable(tidy(lm2))
```

### Interpretation  
When controlling for age, sex, and location:

Free books

Effect becomes 0.0082 (0.82 percentage points)

No longer significant (p = 0.21)

Interpretation: The program likely does not have a true causal effect once demographic factors are accounted for.

Age

Each additional year increases enrollment by 0.00448 (0.45 percentage points)

Statistically significant (p = 0.0167)

Sex (Male = 1)

Boys are 1.50 percentage points less likely to be enrolled

Significant (p = 0.0213)

Urban location

Urban children are 8.94 percentage points more likely to be enrolled

Strongly significant (p < 0.0001)

## 6.3 Graph: Predicted Probabilities

```{r}
rct$predicted <- predict(lm2)

ggplot(rct, aes(x=factor(freebooks), y=predicted, fill=factor(freebooks))) +
  geom_boxplot() +
  scale_x_discrete(labels=c("0"="Control","1"="Free Books")) +
  labs(
    title="Predicted Enrollment from Regression Model",
    x="Group",
    y="Predicted Probability"
  ) +
  theme_minimal()
```

This graph shows differences in predicted enrollment between groups after adjusting for age, sex, and location.

### RCT Policy Note

**Answer:**  
Free books alone do not meaningfully increase school attendance, though they may provide a small benefit. More impactful policies should target rural areas and boys’ enrollment.


