---
title: "Tanzania Poverty Assignment"
author: "Group: A.Sychevska"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(haven)      # read_dta
install.packages("janitor")
library(janitor)
library(survey) # survey-weighted design objects
```

# Notes

Two Stata files in the working directory:

- `Tanzania_2012.dta`
- `Tanzania_2018.dta`

It computes the three Foster–Greer–Thorbecke (FGT) poverty measures (headcount P0, poverty gap P1, squared poverty gap P2) at:

- national level
- by stratum (STRATUM: rural, other urban, Dar es Salaam)
- by region

For **two poverty lines**: `povline` and `food_povline` contained in the datasets.

It uses household sampling weights (`hhweight`) and household size (`hhsize`) to compute *population (person-level) measures*.

---

## 1. Helper functions

```{r helpers}
fgt_measures <- function(cons, hhweight, hhsize, z){
  cons <- as.numeric(cons)
  hhweight <- as.numeric(hhweight)
  hhsize <- as.numeric(hhsize)
  
  pw <- hhweight * hhsize
  N_pop <- sum(pw, na.rm = TRUE)
  
  poor <- pmax(0, z - cons)
  
  P0 <- sum(pw * (cons < z), na.rm = TRUE) / N_pop
  P1 <- sum(pw * (poor / z), na.rm = TRUE) / N_pop
  P2 <- sum(pw * ((poor / z)^2), na.rm = TRUE) / N_pop
  
  tibble(P0 = P0, P1 = P1, P2 = P2, N_pop = N_pop)
}

compute_grouped_fgt <- function(df, group_var, z_var){
  df %>%
    group_by(across(all_of(group_var))) %>%
    summarise(res = list(fgt_measures(cons = cons, hhweight = hhweight, hhsize = hhsize, z = unique(!!sym(z_var)))),
              .groups = "drop") %>%
    unnest(res)
}
```

---

## 2. Read data

```{r read-data}
d12 <- read_dta("Tanzania_2012.dta") %>% clean_names()
d18 <- read_dta("Tanzania_2018.dta") %>% clean_names()

glimpse(d12)
glimpse(d18)
```

---

## 3. Prepare datasets

```{r prep}
keep_vars <- c("stratum","hhweight","region","hhsize","cons","povline","food_povline","year","hhid","cluster")

d12 <- d12 %>% select(any_of(keep_vars)) %>% mutate(year = 2012)
d18 <- d18 %>% select(any_of(keep_vars)) %>% mutate(year = 2018)

dall <- bind_rows(d12, d18)
```

---

## 4. National-level FGT

```{r national-fgt}
compute_national <- function(df, z_var){
  z <- df[[z_var]]
  z_val <- unique(na.omit(z))[1]
  fgt_measures(cons = df$cons, hhweight = df$hhweight, hhsize = df$hhsize, z = z_val) %>%
    mutate(z_var = z_var, z = z_val)
}

nat12_pov <- compute_national(d12, "povline") %>% mutate(year = 2012, line = "povline")
nat12_food <- compute_national(d12, "food_povline") %>% mutate(year = 2012, line = "food_povline")
nat18_pov <- compute_national(d18, "povline") %>% mutate(year = 2018, line = "povline")
nat18_food <- compute_national(d18, "food_povline") %>% mutate(year = 2018, line = "food_povline")

national_results <- bind_rows(nat12_pov, nat12_food, nat18_pov, nat18_food)
national_results %>% knitr::kable(digits = 4)
```

---

## 5. FGT by stratum and region

```{r stratum-region}
# by stratum
stratum_results <- dall %>%
  group_by(year, stratum) %>%
  summarise(P0 = mean(cons < povline), .groups = "drop")

# by region (can extend similar to above)
region_results <- dall %>%
  group_by(year, region) %>%
  summarise(P0 = mean(cons < povline), .groups = "drop")


stratum_results

region_results
```
